Supplementary Material for `The response of trophic interaction networks to multiple stressors in a marine latitudinal gradient of the Southern Hemisphere'

### 2024-06-10 manuscript.tex
Updated manuscript with minor revisions

\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}

@@

Here we review for the first time stressor responses of marine food webs
for areas located along a Southwest Atlantic to Antarctic gradient (45 -
78ºS, Figure 1). We considered proven and expected changes in food webs
driven by global change-related stressors in selected areas along this
large-scale latitudinal range. The aim of this review is threefold: 1)
describe the complexity and structure of marine food webs along the
Southwest Atlantic to Antarctic axis from a network perspective; 2)
identify the ongoing global change-related stressors for each marine
ecosystem containing the food webs; and 3) review proven food web
changes and elaborate on a series of alternative hypotheses on how the
identified stressors might affect food web features (energy flow,
stability and topology), combining information on node- and
network-level properties. To identify the main stressors, we performed
an exploratory and configurative review instead of a systematic review
(Haddaway et al. 2015), because our objective was to formulate
evidenced-based hypotheses rather than providing an exhaustive list of
species-level stressors for each ecosystem. Finally, we suggest which
additional data and analyses are needed to gain insights into the
stressors' effects on food web properties in the Southwest Atlantic to
Antarctic region.

@@

The Southwest and the Atlantic sectors of the Southern Ocean comprise
one of the most biologically productive regions of the world's oceans
(Acha et al. 2004; Latorre et al. 2023). The referred region extends
from San Jorge Gulf (45ºS) in the Patagonian shelf to the Weddell Sea
(78ºS) in the Southern Ocean and covers a well-connected oceanic area
(Matano et al. 2010; Guihou et al. 2020).

Throughout this region, many studies address the ecology of
predator-prey relationships but most of these studies are restricted to
a few interacting species or certain components of the food web such as
fish, crustaceans, and benthos (e.g., Vinuesa and Varisco (2007),
Padovani et al. (2012), Pasotti et al. (2015), Riccialdelli et al.
(2020), Saunders et al. (2019), Pasti et al. (2021), Cordone et al.
(2022)). Only few studies consider the complexity of trophic
interactions within highly resolved food webs containing hundreds of
species, ranging from phytoplankton, zooplankton, benthos, fish,
seabirds to whales (but see Jacob et al. (2011), Marina et al. (2018),
Funes et al. (2022), López-López et al. (2022), Rodriguez et al. (2022),
Marina et al. (in rev.)). Neglecting this complexity might lead to a
misunderstanding of the structure and functioning of the ecosystems and
ultimately reduce the ability to predict ecosystem responses to
perturbations (Montoya et al. 2009).

@@

In this review, we included six large marine areas in the Southern
Hemisphere for which highly-resolved empirical food webs in terms of
species diversity and trophic interactions have been previously
described: (1) San Jorge Gulf (\textasciitilde45 - 47ºS, 65 - 68ºW)
(Funes et al. 2022), (2) Beagle Channel (\textasciitilde54ºS, 68ºW)
(Rodriguez et al. 2022), (3) Burdwood Bank (\textasciitilde54ºS, 59ºW)
(Marina et al. in rev.), (4) Scotia Sea (\textasciitilde48 - 58ºS, 50ºW)
(López-López et al. 2022), (5) Potter Cove (62ºS, 58ºW, Antarctica)
(Marina et al. 2018), and (6) Weddell Sea (74 - 78ºS, 30ºW, Antarctica)
(Jacob et al. 2011; Marina et al. 2024) (Figure 1). The geographic
locations of these marine biomes span from temperate Atlantic waters to
Antarctic, and are exposed to global change stressors (e.g., sea
warming, glacial retreat, fishery, pollution) (Sage 2020), of which some
are locally more important, e.g., fisheries in the San Jorge Gulf or
glacial retreat in the Antarctic.

@@

Located at a more southern latitude (\textasciitilde54ºS, 68ºW), the
Beagle Channel (Figure 1) is an interoceanic passage at the southernmost
tip of South America, spanning 240 km in length and 5 km in width
(\textasciitilde1200 \(\text{km}^2\)), with a depth range of 0 - 140 m.
It features complex coastlines, varying bathymetry, a prevailing
west-to-east circulation pattern, and a significant longitudinal
gradient of glacial freshwater discharge (Schloss et al. 2023). Bruno et
al. (2023) suggests that locally produced suspended particulate organic
matter (mostly composed of phytoplankton) and organic matter accumulated
in the sediments (macroalgae-originated detritus) are the primary food
sources for the marine ecosystem, as opposed to allochthonous materials.

@@

The so-called Burdwood Bank ecosystem comprises Marine Protected Areas
Namuncurá - Burdwood Bank I and II (\textasciitilde54ºS, 59ºW), meaning
the shallow submarine plateau named Burdwood Bank with a 200 m isobath
boundary, and a deep slope that reaches 4000 m in depth, respectively
(Administración de Parques Nacionales 2022) (Figure 1). Physical
features in the plateau are fairly stable, with salinity averaging 34
all year round and temperature ranging between 4 and 8ºC (Acha et al.
2004). The plateau is surrounded by steep flanks of up to 4000 m depth,
protected by the Marine Protected Area Namuncurá - Burdwood Bank II
(32,000 \(\text{km}^2\)). Intense upwelling and mixing occur in relation
to the slope, entraining deep nutrient-rich waters into the photic layer
(Matano et al. 2019), and resulting in a fairly homogeneous water column
both spatially and temporally (Matano et al. 2019).

The Burdwood Bank food web comprises 379 nodes and 1788 interactions,
with a connectance of 0.01, and an asymmetric degree distribution
(i.e.~most of the species have a relatively low number of interactions
and few species concentrate most of them). Almost half of the consumers
are omnivores (48\%), and the network displays a small-world pattern
(Marina et al. in rev.) (Table 1). This network pattern implies a rapid
spread of a perturbation (e.g., contaminant, population fluctuation,
local extinction) throughout the food web due to a short distance among
species but, at the same time, a greater resistance caused by the
confinement of perturbations mainly within subnetworks as a result of a
high clustering coefficient (Bornatowski et al. 2017; Dormann et al.
2017).

@@

The Scotia Sea is a deep-sea basin (\textasciitilde48 - 58ºS, 50ºW),
delimited by the Drake Passage to the West and by the island complex of
the Scotia Arc to the North, East, and South, with an approximate
extension of 1.5 x 106 \(\text{km}^2\) and a depth range of 0 - 3000 m
(Murphy et al. 2006) (Figure 1). Its oceanography is dominated by the
Antarctic Circumpolar Current, which is spatially structured by frontal
systems (Whitworth 1980). The South Antarctic Circumpolar Current Front
subdivides the Scotia Sea into two biogeographic regions: the Northern
Scotia Sea, characterized by higher and more variable temperatures, and
the Southern Scotia Sea, characterized by lower and more stable
temperatures and influenced by seasonal sea ice (Raymond 2011).

@@

Located between 74 and 78ºS, the high Antarctic Weddell Sea shelf spans
approximately 450 km from East to West (Jacob et al. 2011) (Figure 1).
The water depth ranges from 200 to 500 m, with shallower regions being
covered by continental ice that forms the coastline along the eastern
and southern parts of the Weddell Sea. The shelf area is characterized
by a complex three-dimensional benthic habitat, substantial benthic
biomasses, and an intermediate to high diversity (Teixidó et al. 2002).

The Weddell Sea food web exhibits a high level of network complexity,
featuring the greatest number of nodes (490) and trophic interactions
(16041) among the marine food webs of the Southern Hemisphere (Table 1).
Its connectance (0.07), mean trophic level (2.62) and omnivory (51\%)
are intermediate compared to the other food webs (Table 1). Recently,
the interaction strengths of this food web were estimated, revealing the
presence of numerous weak and few strong interactions (Marina et al.
2024).

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1538}}
  >{\centering\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0839}}
  >{\centering\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0839}}
  >{\centering\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1259}}
  >{\centering\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1259}}
  >{\centering\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0979}}
  >{\centering\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1049}}
  >{\centering\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2238}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Food web}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{Nodes}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{Links}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{Connectance}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{Path length}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{mean TL}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{Omnivory}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{Reference}
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
San Jorge Gulf & 165 & 1015 & 0.04 & 2.17 & 3.02 & 0.63 & Funes et al.
(2022) \\
Beagle Channel & 145 & 1115 & 0.05 & 2.12 & 2.37 & 0.55 & Rodriguez et
al. (2022) \\
Burdwood Bank & 379 & 1788 & 0.01 & 2.99 & 2.52 & 0.49 & Marina et al.
(in rev.) \\
Potter Cove & 110 & 649 & 0.05 & 2.33 & 2.22 & 0.46 & Marina et al.
(2018); Rodriguez et al. (2022) \\
N Scotia Sea & 218 & 10008 & 0.21 & 1.87 & 3.29 & 0.73 & López-López et
al. (2022) \\
S Scotia Sea & 192 & 7241 & 0.20 & 1.90 & 3.21 & 0.71 & López-López et
al. (2022) \\
Weddell Sea & 490 & 16041 & 0.07 & 2.19 & 2.62 & 0.51 & Jacob et al.
(2011) \\
\end{longtable}

@@

\bibitem[\citeproctext]{ref-Marina2024}
Marina, T.I., Saravia, L.A., and Kortsch, S. 2024. New insights into the
{Weddell Sea} ecosystem applying a quantitative network approach. Ocean
Science \textbf{20}(1): 141--153. Copernicus GmbH.
doi:\href{https://doi.org/10.5194/os-20-141-2024}{10.5194/os-20-141-2024}.

\bibitem[\citeproctext]{ref-Marinainrev}
Marina, T.I., Schloss, I.R., Lovrich, G.A., Boy, C.C., Bruno, D.O.,
Capitanio, F.L., Delpiani, S.M., Díaz de Astarloa, J.M., Fraysse, C.,
García Alonso, V.A., Raya Rey, A., Schejter, L., Spinelli, M., Tatián,
M., Urteaga, D., and Riccialdelli, L. in rev. The complex network of
trophic interactions in a {Subantarctic} oceanic {Marine Protected
Area}. Marine Ecology Progress Series.

>>>>

\usepackage{bookmark}

@@

Here we review for the first time stressor responses of marine food webs
for areas located along a Southwest Atlantic to Antarctic gradient (45 -
78°S, Figure 1). We considered proven and expected changes in food webs
driven by global change-related stressors in selected areas along this
large-scale latitudinal range. The aim of this review is threefold: 1)
describe the complexity and structure of marine food webs along the
Southwest Atlantic to Antarctic axis from a network perspective; 2)
identify the ongoing global change-related stressors for each marine
ecosystem containing the food webs; and 3) review proven food web
changes and elaborate on a series of alternative hypotheses on how the
identified stressors might affect food web features (energy flow,
stability and topology), combining information on node- and
network-level properties. To identify the main stressors, we performed
an exploratory and configurative review instead of a systematic review
(Haddaway et al. 2015), where we considered more than 200 peer-reviewed 
articles retrieved mainly from Google Scholar using keywords combining 
location ( "San Jorge Gulf", "Beagle Channel", "Burdwood Bank", "Scotia Sea", 
"Weddell Sea", "Potter Cove") and stressors ("acidification", "alien species", 
"climate change", "contamination", "fishery", "glacier retreat", "global warming", 
"ice scouring", "invasive species", "marine contaminant", "microplastics", "mercury", 
"pollution", "sea ice", "sea warming", "urban waste"). Additionally, we 
specifically searched for publications by authors we know may have conducted 
related studies in the area. We performed this type of review because our 
objective was to formulate
evidenced-based hypotheses rather than providing an exhaustive list of
species-level stressors for each ecosystem. Finally, we suggest which
additional data and analyses are needed to gain insights into the
stressors' effects on food web properties in the Southwest Atlantic to
Antarctic region.

@@

The Southwest and the Atlantic sectors of the Southern Ocean comprise
one of the most biologically productive regions of the world's oceans
(Acha et al. 2004; Latorre et al. 2023). The referred region extends
from San Jorge Gulf (45°S) in the Patagonian shelf to the Weddell Sea
(78°S) in the Southern Ocean and covers a well-connected oceanic area
(Matano et al. 2010; Guihou et al. 2020).

Throughout this region, many studies address the ecology of
predator-prey relationships but most of these studies are restricted to
a few interacting species or certain components of the food web such as
fish, crustaceans, and benthos (e.g., Vinuesa and Varisco (2007),
Padovani et al. (2012), Pasotti et al. (2015), Riccialdelli et al.
(2020), Saunders et al. (2019), Pasti et al. (2021), Cordone et al.
(2022)). Only few studies consider the complexity of trophic
interactions within highly resolved food webs containing hundreds of
species, ranging from phytoplankton, zooplankton, benthos, fish,
seabirds to whales (but see Jacob et al. (2011), Marina et al. (2018),
Funes et al. (2022), López-López et al. (2022), Rodriguez et al. (2022),
Marina et al. (2024b)). Neglecting this complexity might lead to a
misunderstanding of the structure and functioning of the ecosystems and
ultimately reduce the ability to predict ecosystem responses to
perturbations (Montoya et al. 2009).

@@

In this review, we included six large marine areas in the Southern
Hemisphere for which highly-resolved empirical food webs in terms of
species diversity and trophic interactions have been previously
described: (1) San Jorge Gulf (\textasciitilde45 - 47°S, 65 - 68°W)
(Funes et al. 2022), (2) Beagle Channel (\textasciitilde54°S, 68°W)
(Rodriguez et al. 2022), (3) Burdwood Bank (\textasciitilde54°S, 59°W)
(Marina et al. 2024b), (4) Scotia Sea (\textasciitilde48 - 58°S, 50°W)
(López-López et al. 2022), (5) Potter Cove (62°S, 58°W, Antarctica)
(Marina et al. 2018), and (6) Weddell Sea (74 - 78°S, 30°W, Antarctica)
(Jacob et al. 2011; Marina et al. 2024a) (Figure 1). The geographic
locations of these marine biomes span from temperate Atlantic waters to
Antarctic, and are exposed to global change stressors (e.g., sea
warming, glacial retreat, fishery, pollution) (Sage 2020), of which some
are locally more important, e.g., fisheries in the San Jorge Gulf or
glacial retreat in the Antarctic.

@@

Located at a more southern latitude (\textasciitilde54°S, 68°W), the
Beagle Channel (Figure 1) is an interoceanic passage at the southernmost
tip of South America, spanning 240 km in length and 5 km in width
(\textasciitilde1200 \(\text{km}^2\)), with a depth range of 0 - 140 m.
It features complex coastlines, varying bathymetry, a prevailing
west-to-east circulation pattern, and a significant longitudinal
gradient of glacial freshwater discharge (Schloss et al. 2023). Bruno et
al. (2023) suggests that locally produced suspended particulate organic
matter (mostly composed of phytoplankton) and organic matter accumulated
in the sediments (macroalgae-originated detritus) are the primary food
sources for the marine ecosystem, as opposed to allochthonous materials.

@@

The so-called Burdwood Bank ecosystem comprises Marine Protected Areas
Namuncurá - Burdwood Bank I and II (\textasciitilde54°S, 59°W), meaning
the shallow submarine plateau named Burdwood Bank with a 200 m isobath
boundary, and a deep slope that reaches 4000 m in depth, respectively
(Administración de Parques Nacionales 2022) (Figure 1). Physical
features in the plateau are fairly stable, with salinity averaging 34
all year round and temperature ranging between 4 and 8°C (Acha et al.
2004). The plateau is surrounded by steep flanks of up to 4000 m depth,
protected by the Marine Protected Area Namuncurá - Burdwood Bank II
(32,000 \(\text{km}^2\)). Intense upwelling and mixing occur in relation
to the slope, entraining deep nutrient-rich waters into the photic layer
(Matano et al. 2019), and resulting in a fairly homogeneous water column
both spatially and temporally (Matano et al. 2019).

The Burdwood Bank food web comprises 379 nodes and 1788 interactions,
with a connectance of 0.01, and an asymmetric degree distribution
(i.e.~most of the species have a relatively low number of interactions
and few species concentrate most of them). Almost half of the consumers
are omnivores (48\%), and the network displays a small-world pattern
(Marina et al. 2024b) (Table 1). This network pattern implies a rapid
spread of a perturbation (e.g., contaminant, population fluctuation,
local extinction) throughout the food web due to a short distance among
species but, at the same time, a greater resistance caused by the
confinement of perturbations mainly within subnetworks as a result of a
high clustering coefficient (Bornatowski et al. 2017; Dormann et al.
2017).

@@

The Scotia Sea is a deep-sea basin (\textasciitilde48 - 58°S, 50°W),
delimited by the Drake Passage to the West and by the island complex of
the Scotia Arc to the North, East, and South, with an approximate
extension of 1.5 x 106 \(\text{km}^2\) and a depth range of 0 - 3000 m
(Murphy et al. 2006) (Figure 1). Its oceanography is dominated by the
Antarctic Circumpolar Current, which is spatially structured by frontal
systems (Whitworth 1980). The South Antarctic Circumpolar Current Front
subdivides the Scotia Sea into two biogeographic regions: the Northern
Scotia Sea, characterized by higher and more variable temperatures, and
the Southern Scotia Sea, characterized by lower and more stable
temperatures and influenced by seasonal sea ice (Raymond 2011).

@@

Located between 74 and 78°S, the high Antarctic Weddell Sea shelf spans
approximately 450 km from East to West (Jacob et al. 2011) (Figure 1).
The water depth ranges from 200 to 500 m, with shallower regions being
covered by continental ice that forms the coastline along the eastern
and southern parts of the Weddell Sea. The shelf area is characterized
by a complex three-dimensional benthic habitat, substantial benthic
biomasses, and an intermediate to high diversity (Teixidó et al. 2002).

The Weddell Sea food web exhibits a high level of network complexity,
featuring the greatest number of nodes (490) and trophic interactions
(16041) among the marine food webs of the Southern Hemisphere (Table 1).
Its connectance (0.07), mean trophic level (2.62) and omnivory (51\%)
are intermediate compared to the other food webs (Table 1). Recently,
the interaction strengths of this food web were estimated, revealing the
presence of numerous weak and few strong interactions (Marina et al.
2024a).

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1538}}
  >{\centering\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0839}}
  >{\centering\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0839}}
  >{\centering\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1259}}
  >{\centering\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1259}}
  >{\centering\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0979}}
  >{\centering\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1049}}
  >{\centering\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2238}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Food web}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{Nodes}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{Links}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{Connectance}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{Path length}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{mean TL}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{Omnivory}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{Reference}
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
San Jorge Gulf & 165 & 1015 & 0.04 & 2.17 & 3.02 & 0.63 & Funes et al.
(2022) \\
Beagle Channel & 145 & 1115 & 0.05 & 2.12 & 2.37 & 0.55 & Rodriguez et
al. (2022) \\
Burdwood Bank & 379 & 1788 & 0.01 & 2.99 & 2.52 & 0.49 & Marina et al.
(2024b) \\
Potter Cove & 110 & 649 & 0.05 & 2.33 & 2.22 & 0.46 & Marina et al.
(2018); Rodriguez et al. (2022) \\
N Scotia Sea & 218 & 10008 & 0.21 & 1.87 & 3.29 & 0.73 & López-López et
al. (2022) \\
S Scotia Sea & 192 & 7241 & 0.20 & 1.90 & 3.21 & 0.71 & López-López et
al. (2022) \\
Weddell Sea & 490 & 16041 & 0.07 & 2.19 & 2.62 & 0.51 & Jacob et al.
(2011) \\
\end{longtable}

@@

\bibitem[\citeproctext]{ref-Marina2024}
Marina, T.I., Saravia, L.A., and Kortsch, S. 2024a. New insights into
the {Weddell Sea} ecosystem applying a quantitative network approach.
Ocean Science \textbf{20}(1): 141--153. Copernicus GmbH.
doi:\href{https://doi.org/10.5194/os-20-141-2024}{10.5194/os-20-141-2024}.

\bibitem[\citeproctext]{ref-Marina2024a}
Marina, T.I., Schloss, I.R., Lovrich, G.A., Boy, C.C., Bruno, D.O.,
Capitanio, F.L., Delpiani, S.M., Díaz De Astarloa, J.M., Fraysse, C.,
García Alonso, V.A., Raya Rey, A., Schejter, L., Spinelli, M., Tatián,
M., Urteaga, D., and Riccialdelli, L. 2024b. The complex network of
trophic interactions in a sub-{Antarctic} oceanic {Marine Protected
Area}. Marine Ecology Progress Series.
doi:\href{https://doi.org/10.3354/meps14600}{10.3354/meps14600}.
